### 2012-08-20 whitepaper/pixel_level.tex
noise models, source sensitivity, blind decon, sampling

\newcommand{\expectation}[1]{\tilde{#1}}
\newcommand{\sky}{^{(\mathrm{sky})}}

@@

\noindent
David W. Hogg (NYU)
\\
\textsl{Max-Planck-Institut f\"ur Astronomie, Heidelberg, Germany}

@@

\paragraph{pixel signal models}
Sticking with the intermediate level of dumbness for now: In a set of
$3\times 3$ sky patches, the expected signal
$\expectation{D}_{ij}$---the expected number of electrons or data
numbers read out---in pixel $i$ of exposure $j$ might be a combination
of zero-level, dark-current, and sky-photon signals
\begin{equation}
\expectation{D}_{ij} = z_i + d_i\,t_j + f\sky_i\,S\sky_j\,t_j
\end{equation}
where $i$ labels the device pixel ($i$ maps back to a pixel location
in the original imaging device), $j$ labels the exposure, each of
which might have a slightly different sky level, $z_i$ is a zero-level
or bias for pixel $i$, $d_i$ is a dark current rate or signal per time
in the absence of illumination, $t_j$ is the exposure time for
exposure $j$, $f\sky_i$ is the sensitivity of pixel $i$ to diffuse sky
illumination, and $S\sky_j$ is the intensity of the sky in exposure
$j$ (interpolated to pixel $i$ if there are gradients).

In a $3\times 3$ source patch, the expectation is similar...
\begin{equation}
\expectation{D}_{ij} = z_i + d_i\,t_j + f\sky_i\,S\sky_j\,t_j
\end{equation}

>>>>

\newcommand{\expectation}[1]{\tilde{#1}}
\newcommand{\sky}{^{(\mathrm{sky})}}
\newcommand{\source}{^{(\mathrm{src})}}

@@

\noindent
David W. Hogg (NYU)
\\{\footnotesize
\textsl{Center for Cosmology and Particle Physics, Department of Physics, New York University, New York, USA}
\\
\textsl{Max-Planck-Institut f\"ur Astronomie, Heidelberg, Germany}}

\textsl{This document is a draft dated 2012-08-20.  It is based in
  part on conversations with Rob Fergus (NYU), Rory Holmes (MPIA), and
  Bernhard Sch\"olkopf (MPI-IS).}

@@

\paragraph{pixel signal models:}
Sticking with the intermediate level of dumbness for now: In a
$3\times 3$ sky patch (a patch far from any detectable sources), the
expected signal $\expectation{D}_{ij}$---the expected number of
electrons or data numbers read out---in pixel $i$ of exposure $j$
might be a combination of zero-level, dark-current, and sky-photon
signals
\begin{eqnarray}
\expectation{D}_{ij} = z_i + d_i\,t_j + f\sky_i\,S\sky_j\,t_j
\quad ,
\end{eqnarray}
where $i$ labels the device pixel ($i$ maps back to a pixel location
in the original imaging device), $j$ labels the exposure, each of
which might have a slightly different sky level, $z_i$ is a zero-level
or bias for pixel $i$, $d_i$ is a dark current rate or signal per time
in the absence of illumination, $t_j$ is the exposure time for
exposure $j$, $f\sky_i$ is the sensitivity of pixel $i$ to diffuse sky
illumination, and $S\sky_j$ is the intensity of the sky in exposure
$j$ (interpolated to pixel $i$ if there are gradients).

In a $3\times 3$ source patch (a patch containing some detectable
source or part thereof), the expectation is similar but there is now
source flux
\begin{eqnarray}
\expectation{D}_{ij} = z_i + d_i\,t_j + f\sky_i\,S\sky_j\,t_j
                     + f\source_i\,S\source_{ij}\,t_j
\quad ,
\end{eqnarray}
where now $f\source_i$ is the sensitivity of pixel $i$ to compact
source flux, and $S\source_{ij}$ is the source flux in exposure $j$
interpolated to or evaluated at the location of pixel $i$.  This
interpolation to pixel $i$ is where the $3\times 3$ pixel patch comes
in: We are going to use the 8 pixels surrounding pixel $i$ to
interpolate the source flux probabilistically to pixel $i$.  This also
shows that using a bigger patch, say $5\times 5$, would be
substantially less dumb.  Even the sky patches (as noted momentarily
above) might require interpolation to pixel $i$, since there might be
sky gradients.

In the case that there are also \emph{dark}
observations---observations with the shutter closed---or \emph{zero}
observations---observations with zero exposure times---there can be
additional constraints on the $z_i$ and the $d_i$ beyond just the
science-data constraints.  These considerations also suggest that a
survey that varies its exposure time will thereby obtain much more
information about the calibration information (and also vet the
shutter!).

There might be an additional dependence of the source sensitivity
$f\source_i$ on color or spectral energy distribution.  There might
also be variations of all the calibration parameters on time, possibly
even with break-points or sharp transitions (think episodic radiation
damage and the like).  There will also be many, many pixels affected
by cosmic rays.  All of these issues make the real-world execution of
this program more complicated than the cartoon given here.  For now we
might try assuming that we can work with only patches that are
obviously \emph{not} affected by cosmic rays, and work only with
time-slices of the data within which no significant variations are
expected.

The likelihood of any model for the data will be related to the idea
that the observed or read-out data $D_{ij}$ differs from the model
expectation $\expectation{D}_{ij}$ by the contribution of some kind of
noise, which might be Poisson or Gaussian or some combination,
depending on the device and our precision needs.

\paragraph{blind deconvolution:}
This problem looks very much like a blind deconvolution problem,
because there is a true but unknown scene multiplied by some unknown
set of pixel responses before noisy detection.  If the astronomical
fields $(S\sky_j, S\source_{ij})$ are known, the pixel response
parameters $(z_i, d_i, f\sky_i, f\source_i)$ are related to the signal
by a pure linear operation and \foreign{vice versa}.  If the noise in
the data is Gaussian (or can be approximated as such), the inference
can proceed by iterated weighted linear least squares.  If the imager
is well-behaved, $S\sky_j$ will be a very smooth function of pixel
position $i$ or no function at all.  If the imaging is well-sampled,
$S\source_{ij}$ will be expressable as a linear combination of
band-limited basis functions, like sincs or Gaussians.

If we aren't satisfied with point estimates---and we shouldn't
be---then we can propagate the uncertainties in both the calibration
quantities and the astronomical scene as a set of samples or explicit
analytic descriptions of the posterior PDF.  The latter might be
possible in the case of pure Gaussian noise.
